Simple repression draft

### 2020-05-21 doc/section_03_beyond_means.tex
RP Beyond Means Thursday
I have finished a first pass through beyond the Means.

\begin{figure}%[h!]
\centering
\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
\caption{\textbf{Constitutive promoter cartoon comparison.}
The left column depicts various plausible cartoons for constitutive promoters.
We also list literature references where models have been used previously. In
model (1), transcripts are produced in a Poisson process~\cite{Jones2014}. Model
(2) features explicit modeling of RNAP binding/unbinding
kinetics~\cite{Phillips2015a}. Model (3) is a more detailed generalization of
model (2), treating transcription initiation as a multi-step process proceeding
through closed and open complexes~\cite{Mitarai2015}. Model (4) is somewhat
analogous to (2) except with the precise nature of active and inactive states
left ambiguous~\cite{Razo-Mejia2020}. Finally, model (5) can be viewed as a
certain limit of model (4) in which transcripts are produced in bursts, and
initiation of bursts is a Poisson process. \mmnote{Haven't found model 5
formulated quite like this in the literature, usually burstiness is handled like
model 4, but I should probably search a bit more.} The right column shows the
Fano factor for each model. Note especially the crucial diagnostic: (2) and (3)
have $\nu$ strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
Models with Fano factors $\le 1$ cannot produce the FISH data observed
in~\cite{Jones2014} without introducing additional assumptions and model
complexity.}
\label{fig:constit_cartoons}
\end{figure}

@@

\subsubsection{Model 2 - Two-state promoter, RNAP bound or unbound.}
This model was considered in, e.g.,~\cite{Phillips2015a} and
~\cite{Phillips2019}. Here transcription initiation proceeds from the bound to
the unbound state, reflecting the microscopic reality that an RNAP that has
begun to elongate a transcript is no longer available at the start site to begin
another. The problem with this picture is that the Fano factor is
\begin{align}
    \nu = 1 -
        \frac{r\rate{k}{P}{+}}
            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)},
\end{align}
which is always $<1$. To make contact with the experimental reality of $\nu>1$,
we will have to do, at a minimum, the same corrections to Poisson behavior as in
model 1 above. So while this model adds an appealing element of microscopic
reality, we are forced to reject it as the additional complexity is unable to
capture the phenomenology of interest. Obviously the promoter state does in fact
proceed through cycles of RNAP binding, initiating, and elongating, but it seems
that the super-Poissonian noise in mRNA copy number we want to model must be
governed by other features of the system.

\subsubsection{Model 3 - Three-state promoter, multistep transcription
initiation and escape.} How might we remedy the deficits of model 2? It is
known~\cite{DeHaseth1998} that once RNAP initially binds the promoter region, a
multitude of distinct steps occur sequentially before RNAP finally escapes into
the elongation phase. Perhaps adding some of this mechanistic detail might
rescue model 2? The next simplest refinement of model 2 could consider open
complex formation and promoter escape as separate steps rather than as a single
effective step. In other words, we construct model 3 by adding a single extra
state to model 2, and we will label the two RNAP-bound states as the closed and
open complexes, despite the true biochemical details certainly being more
complex. The authors of~\cite{Mitarai2015}, for instance, considered this model,
although they added an additional repressor bound state and did not explicitly
consider the limit with no repressor that we analyze here. Again, our goal here
is not a complete accounting of all the relevant biochemical detail; this is an
exploratory search for the important features our model needs to include.

Unfortunately, as the authors of~\cite{Mitarai2015} hint, this model too has
Fano factor $\nu<1$. This can be seen by starting with their Eq.~A1 for the Fano
factor of the analogous model with repressor and taking the limit as repressor
concentration goes to zero. After substantial algebra, the result is
\begin{align}
\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
\frac{k_P^+ + k_P^- + k_O + \gamma}
        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
\end{align}
where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
tidiness. This is necessarily less than 1 for arbitrary rate constants.

In fact, we suspect \textit{any} model in which transcription proceeds through a
multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
the waiting time distribution to traverse the cycle with the waiting time for a
Poisson promoter (model 1) with the same mean time. The latter is simply an
exponential distribution. The former is a convolution of multiple exponentials,
and intuitively should be more peaked with a smaller fractional width than an
exponential with the same mean.\footnote{This can be made more precise with a
result from~\cite{Stewart2007}, who showed that the convolution of multiple
gamma distributions (of which the exponential distribution is a special case) is
to a very good approximation also gamma distributed. Using their Eq.~2 for the
distribution of the convolution, with shape parameters set to 1 to give
exponential distributions, the total waiting time distribution has a ratio of
variance to squared mean $\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i
k_i\right)^2$, where the $k_i$ are the rates of the individual steps. Clearly
this is less than 1 and therefore the total waiting time distribution is
narrower than the corresponding exponential.} A less disperse waiting time
distribution means transcription initations are more uniformly distributed in
time relative to a Poisson process. Hence the distribution of mRNA over a
population of cells should be less variable compared to Poisson, giving
$\nu<1$.\footnote{This last step, while intuitive, can be argued by analogy to
photon statistics where antibunching gives rise to sub-Poissonian
noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do not expect they
apply for our problem. Nevertheless we refrain from elevating this
cycle/sub-Poissonian equivalence to a ``theorem.'' \mmnote{Antibunching is the
``obvious'' analogy for me but it'd be totally out of left field for most
readers, any suggestions on a more comprehensible reference??}}

Regardless of the merits of this model in describing the noise properties of
constitutive transcription initation, it does not capture the dominant source of
noise we want to include in our phenomenological model, so we must discard these
details and search elsewhere.

\subsubsection{Model 4 - Two-state promoter, ``active'' and ``inactive''
states.} Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model 2, but
the interpretation of the two states is changed. Rather than explicitly viewing
them as RNAP bound and unbound, we view them as ``active'' and ``inactive,''
which are able and unable to initiate transcripts, respectively. We are
noncommittal as to the microscopic details of these states.

One interpretation~\cite{Chong2014, Sevier2016} for the active and inactive
states is that it represents the promoter's supercoiling state: transitions to
the inactive state are caused by accumulation of positive supercoiling, which
inhibits transcription, and transitions back to ``active'' are caused by gyrase
or other topoisomerases relieving the supercoiling. This is an interesting
possibility because it would mean the timescale for promoter state transitions
is driven by topoisomerase kinetics, not by RNAP kinetics. From in vitro
measurments, the former are suggested to be of order minutes~\cite{Chong2014}.
Contrast this with model 2, where the state transitions are assumed to be
governed by RNAP, which, assuming a copy number per cell of order $10^3$, has a
diffusion-limited association rate $k_{on} \sim 10^2~\text{s}^{-1}$ to a target
promoter. Combined with known $K_D$'s of order $\mu$M, this gives an RNAP
dissociation rate $k_{off}$ of order $10^2$. As we will show below, however,
there are some lingering puzzles with interpreting this supercoiling
interpretation, so we leave it as a speculation and refrain from assigning
definite physical meaning to the two states in this model.

Intuitively one might expect that, since transcripts are produced as a Poisson
process only when the promoter is in one of the two states in this model,
transcription initiations should now be ``bunched,'' in constrast to the
``anti-bunching'' of models 2 and 3 above. One might further guess that this
bunching would lead to super-Poissonian noise in the mRNA distribution over a
population of cells. An honest calculation of the Fano factor produces
\begin{align}
\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
\end{align}
which is strictly greater than 1, verifying the above intuition. Note we have
dropped the $P$ label on the promoter switching rates to emphasize that these
very likely do not represent kinetics of RNAP itself. This calculation can also
be sidestepped by noting that the model is mathematically equivalent to the
simple repression repression model from~\cite{Jones2014}, with states and rates
relabeled and reinterpreted.

How does this model compare to model 1 above? In model 1, all non-Poisson
features of the mRNA distribution were handled as extrinsic corrections. By
contrast, here the 3 parameter model is used to fit the full mRNA distribution
as measured in mRNA FISH experiments. In essense, all variability in the mRNA
distribution is regarded as ``intrinsic,'' arising either from stochastic
initiation or from switching between the two coarse-grained promoter states. The
advantage of this approach is that it fits neatly into the master equation
picture, and the parameters thus inferred can be used as input for more
complicated models with regulation by transcription factors.

@@

\subsubsection{Model 5 - One-state promoter with explicit bursts.}
This model is inspired by the failure mode of model 4. The key observation above
was that, as found in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
ratio $r/k^-$, could be directly inferred from the FISH data
from~\cite{Jones2014}. So let us take this seriously and imagine a model where
these are the only two model parameters. What would this model look like?

To develop some intuition, consider model 4 in the limit $k^+ \ll k^- \lesssim
r$, which is roughly satisfied by the parameters inferred
in~\cite{Razo-Mejia2020}. In this limit, the system spends the majority of its
time in the inactive state, occasionally becoming active and making a burst of
transcripts. This should call to mind the well-known phenomenon of
transcriptional bursting, as reported in,
e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some
more cites here}. Let us make this correspondence more precise. The mean dwell
time in the active state is $1/k^-$. While in this state, transcripts are
produced at a rate $r$ per unit time. So on average, $r/k^-$ transcripts are
produced before the system switches to the inactive state. Once in the inactive
state, the system dwells there for an average time $1/k^+$ before returning to
the active state and repeating the process. $r/k^-$ resembles an average burst
size, and $1/k^+$ resembles the time interval between burst events. More
precisely, $1/k^+$ is the mean time between the end of one burst and the start
of the next, whereas $1/k^+ + 1/k^-$ would be the mean interval between the
start of two successive burst events, but in the limit $k^+ \ll k^-$, $1/k^+ +
1/k^- \approx 1/k^+$. Note that this limit ensures that the waiting time between
bursts is approximately exponentially distributed, with mean set by the only
timescale left in the problem, $1/k^+$.\footnote{If instead it were the case
that $k^+ \sim k^-$, then the waiting time $1/k^+ + 1/k^-$ would have a peaked
distribution, but this does not appear to be the case for any of datasets
from~\cite{Jones2014}.}

>>>>

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig02.pdf}
\caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
The left column depicts various plausible models for the dynamics of constitutive promoters.
%We also list literature references where models have been used previously. 
In
model (1), transcripts are produced in a Poisson process~\cite{Jones2014}. Model
(2) features explicit modeling of RNAP binding/unbinding
kinetics~\cite{Phillips2015a}. Model (3) is a more detailed generalization of
model (2), treating transcription initiation as a multi-step process proceeding
through closed and open complexes~\cite{Mitarai2015}. Model (4) is somewhat
analogous to (2) except with the precise nature of active and inactive states
left ambiguous~\cite{Razo-Mejia2020}. Finally, model (5) can be viewed as a
certain limit of model (4) in which transcripts are produced in bursts, and
initiation of bursts is a Poisson process. 
%\mmnote{Haven't found model 5
%formulated quite like this in the literature, usually burstiness is handled like
%model 4, but I should probably search a bit more.} 
The right column shows the
Fano factor for each model. Note especially the crucial diagnostic: (2) and (3)
have $\nu$ strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
Models with Fano factors $\le 1$ cannot produce the FISH data observed
in~\cite{Jones2014} without introducing additional assumptions and model
complexity.}
\label{fig:constit_cartoons}
\end{figure}

@@

\subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}
Our second model of constitutive transcription posits an architecture
in which the promoter is either empty or bound by RNAP
~\cite{Phillips2015a, Phillips2019}. Here,  as shown in 
the second row of Figure~\ref{fig:constit_cartoons},  transcription initiation results in
a state transition from the bound to
the unbound state, reflecting the microscopic reality that an RNAP that has
begun to elongate a transcript is no longer available at the start site to begin
another.  As shown in the appendix, the Fano factor in this model is given by
\begin{align}
    \nu = 1 -
        \frac{r\rate{k}{P}{+}}
            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
\end{align}
The problem with this picture is that the Fano factor is
 always $<1$. 
 To make contact with the experimental reality of $\nu>1$,
clearly some corrections will be needed.
While this model adds an appealing element of microscopic
reality, we are forced to reject it as the additional complexity is unable to
capture the phenomenology of interest. Obviously the promoter state does in fact
proceed through cycles of RNAP binding, initiating, and elongating, but it seems
that the super-Poissonian noise in mRNA copy number we want to model must be
governed by other features of the system.

\subsubsection{Noise in the Three-State Promoter, Multistep Transcription
Initiation and Escape.} How might we remedy the deficits of model 2? It is
known~\cite{DeHaseth1998} that once RNAP initially binds the promoter region, a
multitude of distinct steps occur sequentially before RNAP finally escapes into
the elongation phase. Perhaps adding some of this mechanistic detail as shown
in Figure~\ref{fig:constit_cartoons} might
rescue the previous model   The next simplest refinement of that model could consider open
complex formation and promoter escape as separate steps rather than as a single
effective step. In other words, we construct model 3 by adding a single extra
state to model 2, and we will label the two RNAP-bound states as the closed and
open complexes, despite the true biochemical details certainly being more
complex.  For example, earlier work extended this model by adding an additional
repressor bound state and did not explicitly consider the limit with no repressor
that we analyze here~\cite{Mitarai2015}.  Again, our goal here
is not a complete accounting of all the relevant biochemical detail; this is an
exploratory search for the important features that a model needs to include
to square with the known experimental reality of constitutive expression.

Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too has
Fano factor $\nu<1$. This can be seen by starting with their Eq.~A1 for the Fano
factor of the analogous model with repressor and taking the limit as repressor
concentration goes to zero. After substantial algebra, the result is
\begin{align}
\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
\frac{k_P^+ + k_P^- + k_O + \gamma}
        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
\end{align}
where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
tidiness. This is necessarily less than 1 for arbitrary rate constants.

\marginpar{\it RP: very fun next paragraph - I like the argument, need to think more}
In fact, we suspect \textit{any} model in which transcription proceeds through a
multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
the waiting time distribution to traverse the cycle with the waiting time for a
Poisson promoter (model 1) with the same mean time. The latter is simply an
exponential distribution. The former is a convolution of multiple exponentials,
and intuitively should be more peaked with a smaller fractional width than an
exponential with the same mean.\footnote{This can be made more precise with a
result from~\cite{Stewart2007}, who showed that the convolution of multiple
gamma distributions (of which the exponential distribution is a special case) is
to a very good approximation also gamma distributed. Using their Eq.~2 for the
distribution of the convolution, with shape parameters set to 1 to give
exponential distributions, the total waiting time distribution has a ratio of
variance to squared mean $\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i
k_i\right)^2$, where the $k_i$ are the rates of the individual steps. Clearly
this is less than 1 and therefore the total waiting time distribution is
narrower than the corresponding exponential.} A less disperse waiting time
distribution means transcription initations are more uniformly distributed in
time relative to a Poisson process. Hence the distribution of mRNA over a
population of cells should be less variable compared to Poisson, giving
$\nu<1$.\footnote{This last step, while intuitive, can be argued by analogy to
photon statistics where antibunching gives rise to sub-Poissonian
noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do not expect they
apply for our problem. Nevertheless we refrain from elevating this
cycle/sub-Poissonian equivalence to a ``theorem.'' \mmnote{Antibunching is the
``obvious'' analogy for me but it'd be totally out of left field for most
readers, any suggestions on a more comprehensible reference??}}

Regardless of the merits of this model in describing the noise properties of
constitutive transcription initiation, it ultimately fails the simplest quantitative
feature of the data, namely that the Fano factor $> 1$ and hence  we must discard this
mechanistic picture
 and search elsewhere.

\subsubsection{Noise in a Two-State Promoter with ``Active'' and ``Inactive''
States}
 Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model 2, but as with
 the models considered in the section on mean expression,
the interpretation of the two states is changed. Rather than explicitly viewing
them as RNAP bound and unbound, we view them as ``active'' and ``inactive,''
which are able and unable to initiate transcripts, respectively. We are
noncommittal as to the microscopic details of these states.

One interpretation~\cite{Chong2014, Sevier2016} for the active and inactive
states is that they represent the promoter's supercoiling state: transitions to
the inactive state are caused by accumulation of positive supercoiling, which
inhibits transcription, and transitions back to ``active'' are caused by gyrase
or other topoisomerases relieving the supercoiling. This is an interesting
possibility because it would mean the timescale for promoter state transitions
is driven by topoisomerase kinetics, not by RNAP kinetics. From in vitro
measurements, the former are suggested to be of order minutes~\cite{Chong2014}.
Contrast this with model 2, where the state transitions are assumed to be
governed by RNAP, which, assuming a copy number per cell of order $10^3$, has a
diffusion-limited association rate $k_{on} \sim 10^2~\text{s}^{-1}$ to a target
promoter. Combined with known $K_d$'s of order $\mu$M, this gives an RNAP
dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$. As we will show below, however,
there are some lingering puzzles with interpreting this supercoiling
hypothesis, so we leave it as a speculation and refrain from assigning
definite physical meaning to the two states in this model.

Intuitively one might expect that, since transcripts are produced as a Poisson
process only when the promoter is in one of the two states in this model,
transcription initiations should now be ``bunched,'' in constrast to the
``anti-bunching'' of models 2 and 3 above. One might further guess that this
bunching would lead to super-Poissonian noise in the mRNA distribution over a
population of cells.  Indeed, as shown in the appendix, a calculation of the Fano factor produces
\begin{align}
\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
\end{align}
which is strictly greater than 1, verifying the above intuition. Note we have
dropped the $P$ label on the promoter switching rates to emphasize that these
very likely do not represent kinetics of RNAP itself. This calculation can also
be sidestepped by noting that the model is mathematically equivalent to the
simple repression model from~\cite{Jones2014}, with states and rates
relabeled and reinterpreted.

How does this model compare to model 1 above? In model 1, all non-Poisson
features of the mRNA distribution were handled as extrinsic corrections. By
contrast, here the 3 parameter model is used to fit the full mRNA distribution
as measured in mRNA FISH experiments. In essence, all variability in the mRNA
distribution is regarded as ``intrinsic,'' arising either from stochastic
initiation or from switching between the two coarse-grained promoter states. The
advantage of this approach is that it fits neatly into the master equation
picture, and the parameters thus inferred can be used as input for more
complicated models with regulation by transcription factors.

@@

\subsubsection{Noise Model for One-State Promoter with Explicit Bursts}
The final model we consider is 
 inspired by the failure mode of model 4. The key observation above
was that, as found in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
ratio $r/k^-$, could be directly inferred from the FISH data
from~\cite{Jones2014}. So let us take this seriously and imagine a model where
these are the only two model parameters. What would this model look like?

\marginpar{\it RP: best description I have ever seen of this.  Good job}
To develop some intuition, consider model 4 in the limit $k^+ \ll k^- \lesssim
r$, which is roughly satisfied by the parameters inferred
in~\cite{Razo-Mejia2020}. In this limit, the system spends the majority of its
time in the inactive state, occasionally becoming active and making a burst of
transcripts. This should call to mind the well-known phenomenon of
transcriptional bursting, as reported in,
e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some
more cites here}. Let us make this correspondence more precise. The mean dwell
time in the active state is $1/k^-$. While in this state, transcripts are
produced at a rate $r$ per unit time. So on average, $r/k^-$ transcripts are
produced before the system switches to the inactive state. Once in the inactive
state, the system dwells there for an average time $1/k^+$ before returning to
the active state and repeating the process. $r/k^-$ resembles an average burst
size, and $1/k^+$ resembles the time interval between burst events. More
precisely, $1/k^+$ is the mean time between the end of one burst and the start
of the next, whereas $1/k^+ + 1/k^-$ would be the mean interval between the
start of two successive burst events, but in the limit $k^+ \ll k^-$, $1/k^+ +
1/k^- \approx 1/k^+$. Note that this limit ensures that the waiting time between
bursts is approximately exponentially distributed, with mean set by the only
timescale left in the problem, $1/k^+$.\footnote{If instead it were the case
that $k^+ \sim k^-$, then the waiting time $1/k^+ + 1/k^-$ would have a peaked
distribution, but this does not appear to be the case for any of datasets
from~\cite{Jones2014}.}
